---
title: "Data Simulation"
date: "Simulation-based settings"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Article:** Lizbeth Naranjo, Carlos J. Perez, Daniel F. Merino (2025). 
A data ensemble-based approach for detecting vocal disorders using replicated acoustic biomarkers from electroglottography.  
*Sensing and Bio-Sensing Research Journal*, vol, num, pages. 

# Simulation-based settings 

The generation process for this three-class classification problem is adapted to introduce replications in the explanatory variables. 
The covariates $x_{ikj}$ are generated from the algorithm 3, considering $n = 300$ subjects, $i = 1,\ldots,n$, $K = 21$ predictor variables,  $k = 1,\ldots,K$, and $J = 3$ replicated measures, $j = 1,\ldots, J$. 
Note that Algorithm 3 simulates the same number of subjects for each class or category $C$, this number is fixed for each class. 
This procedure enables the simulation of a dataset with many covariates and replicated measures. 

```{r, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE, cache=FALSE, results=TRUE,  comment=""}   
## change the address where the file will be saved 
address = "~/Documents/GitHub/" 
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
set.seed(12345) ## seed 
G = 3   ## number of classes or groups 
n = 300   ## sample size 
m = n/G   ## n/G = sample size / number of classes 
K = 21   ## predictor variables 
J = 3   ## number of replicated measures 

## shifted triangular waveforms functions: 
v1 <- function(k){ max(6-abs(k-11),0) }
v2 <- function(k){ v1(k-4) }
v3 <- function(k){ v1(k+4) }

## function for predictors variables with replications 
predictors <- function(K,G){ 
  u = runif(1) 
  eps = rnorm(K)  
  x = matrix(NA,G,K)
  for(k in 1:K){ 
    x[1,k] = u*v1(k) + (1-u)*v2(k) + eps[k]   ## predictors for class 1
    x[2,k] = u*v1(k) + (1-u)*v3(k) + eps[k]   ## predictors for class 2
    x[3,k] = u*v2(k) + (1-u)*v3(k) + eps[k]   ## predictors for class 3
  } 
  y = c(1,2,3)   ## response variable 
  return(list(x=x,y=y)) 
}

Ytrue = matrix(NA,n)   ## response variable 
X = array(NA,dim=c(n*J,K))   ## K explanatory variables with J replications 
colnames(X) = paste0("V",c(1:K))
ID = rep(NA,n*J)   ## ID of the subject 
Rep = rep(NA,n*J)   ## ID for the replication  

aux = 0
for(i in 1:m){
  for(j in 1:J){
    train <- predictors(K,G)   ## generate variables with replications 
    for(h in 1:G){
      aux = aux+1
      Ytrue[aux] = train$y[h]   
      X[aux,] = train$x[h,]  
      ID[aux] = G*(i-1)+h  
      Rep[aux] = j   
} } }
Y = factor(Ytrue)   ## categorical response variable 
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
pairs(X[,1:7], col=Ytrue,cex=0.5)
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
pairs(X[,8:14], col=Ytrue,cex=0.5)
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
pairs(X[,15:21], col=Ytrue,cex=0.5)
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
## Save the data 
data = data.frame(X,ID,rep=Rep, status=Y)
head(data)
write.table(data,paste0(address,"data_simulated.csv"),row.names=FALSE,sep=";",dec=".")
```
